Efficient inference, potential, and limitations of site-specific substitution models

### 2020-07-26 manuscript/response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\color{black}\textbf{Response:} #1}}

\begin{document}
\section*{Response to reviewers}

We have revised our manuscript extensively in response to the reviewer comments.
In particular, we have put more emphasis on the parts of the manuscript that address estimation of site-specific preferences as means to learn about the biology of the organism as a opposed tto nuisance parameters in phylogenetic inference.
We now also stress that we are not studying site-heterogeneity in the form of a small number of data partitions, but are interested in the ability to estimate preferences for every site in the alignment.
Overall, we feel that the manuscript has much improved from the this refocus and has become simpler to follow.

\editor{This seems like a good paper but in its current state virus evolution may not be a very good fit for it.  I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.}

\response{
Thank you very much for handling the paper and providing these two reviews.
We realized that our original manuscript was misleading to people familar complex mixture models that pratition the seqeunce data into categories with similar properties.
We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specifci properties of every columns of an alignment.
Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available.

}

@@

\refa{- For the simulations, the author draw site-specific preferences from a Dirichlet distribution with $\alpha$ equal to 0.2 or 0.5. It would be nice to present a more interpretable value of the preferences as well, such as entropy or effective number.}

\response{
Thank you for this suggestion.
For amino acid alphabets, $\alpha=0.2$ and $\alpha=0.5$ correspond to an average entropy of $1.8$ or $2.3$ (in natural log units), respectively.
Calculating the effective number of states as $\left(\sum_i p_i^2\right)^{-1}$ yields 4.7 or 7.8 effective states.
We added this to the manuscript.}

\refa{- Fig 5 and Fig S4 are both over-plotted and making the points transparent or adding histograms would help with interpretability. In Fig 1, the differences in linetype should be included in the legend. In Fig 2, it would be easier to keep the models straight if the lines were annotated on the graph directly instead of relying exclusively on color.}

\response{
We are grateful for these suggestions to improve the figures:
\begin{itemize}
    \item we changed Fig 5 and Fig S4 to two dimensional histograms
    \item we added second legend to figure 1 that maps line style to samples size. We also removed the curves corresponding to naive estimates by alignment frequencies. These are also include in Fig 2 and hence not needed here.
    \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
\end{itemize}
}

@@

\subsection*{Reviewer: 2}

\refb{Abstract, line 19: I would phrase this more carefully, as I personally feel that site-specific substitution models are in fact quite popular (with indeed their own site-specific evolutionary rate being assigned/assumed as well) and easy to set up in many inference packages.}

\response{Thank you for this note. We realized that the intention and model class studied in our manuscript wasn't clear enough. We now emphasize that we study whether models with different state preferences at every site can be accurately estimated.

We now point out in multiple places that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
}

@@

\refb{Abstract, line 22: why ‘approximate’? How does this differ from ‘regular’ ML?}

\response{The scheme is approximate, in that the parameters that maximize the likelihood are found through iterative solution of extremal conditions obtained through an approximation of the full likelihood. }

\refb{Abstract, lines 24-25: I assume it will be pointed out that the BEAST/BEAST2 packages do allow for site-specific models to be used (with for example a simple button that allows to select the SRD06 model, of which the paper should be cited)?}

\response{We now discuss site-heterogenous partition models carefully in the main text.}

@@

\refb{Lines 120-122: This brings to mind a recent paper by Gascuel and Steel ‘A Darwinian Uncertainty Principle’ in which the issue of jointly estimating the ancestral states and the transitions of those states is investigated. Given the parameters the authors are interested in here and their assessment through simulations, it seems to me that the authors need to clarify the novelty of their investigations and how theirs sets them apart from those investigated by Gascuel and Steel (actually Figure 1 reminds me of Figure 2 in the paper by Gascuel and Steel).}

\response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do. Gascuel and Steel investigate whether the state of the root if a character can be inferred jointly with the equilibrium probabilities or the transition matrix. Consistent with our results, they find that this is possible for Yule trees.
Here, we are mostly concerned with estimating the model, not the root state, and treat ancestral states as nuissance parameters that at least in principle can be integrated out.
In constrast to Gascuel and Steel, we consider sequence evolution rather than the evolution of a single character which allows our model to share the symmetric transmission matrix across sites.
It also allows us to investigate the extent to which branch lengths can be estimated along with the model.

@@

We now discuss our results in the light of the results by Gascuel and Steel and thank the reviewer for pointing out this work to us.
}

\refb{Line 141: True, and this deserves a lot more explanation in the manuscript, i.e. how far should one typically push the site-specific assumptions? The authors need to make it explicitly clear if they plan to estimate (and assess) one or more site-specific parameters for each site, or whether some of these parameters are shared across categories of sites as is most commonly done when dealing with data partitions (such as partitioning by codon position). If each site gets its own parameter(s), then overfitting seems to be a very real issue; but what about other cases such as pre-defined partitions? Or what about drawing certain parameters from a certain distribution or a hierarchical phylogenetic model, which could very much aid with overfitting issues by sharing data?}

\response{We indeed investigate models with multiple free parameters per site -- in contrast to category models.
In our model the symmetric substitution matrix is shared across sites, but each site has has its own rate and frequencies.
Much of the remainder of the manuscript is devoted to the issue of identifiablility and over-fitting.
As data sets grow in size, such site specific parameters become increasingly identifiable and this is focus of this study.
The properties of category models have for example been studied systematically by Kainer and Lanfear.

We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of model inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
}

@@

\refb{Line 157: 'tree'}
\response{fixed, thanks.}

\refb{Lines 204-208: Again, this seems related to Gascuel and Steel.}

\response{Indeed, Gascuel and Steel show a similar graph for the joint estimation of ancestral state of a single discrete character and the model of its evolution. We now point out this analogy.

The connection to Gascuel and Steel is particularly apt when discussion the inability to jointly infer the ancestral state and the model.
We also draw attention to the fact that accurate model estimation is possible for known trees, consistent with the results by G\&S that joint ancestral reconstruction and model inference is possible for Yule trees.
}

@@

\refb{Lines 239-240: I’m not particularly convinced that uniform frequencies are still the norm ...}

\response{We are not making a statement here on what is the norm and we are aware that complex models with multiple partitions are common. However, many large phylogenetic analysis of viruses use models that account for rate heterogeneity among sites, but use a common matrix of relative substitution rates for all site (PMID 25278604 for example). }

\refb{Line 308: I fear this example analysis is quite (or too) technical and am not sure this will appeal to typical readers of Virus Evolution. Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}

\response{We have made an effort to make this section more accessible, but we think it is an important part of the manuscript and one of the reasons we think Virus Evolution is an appropriate journal.
This section explores the ability to estimate site specific preferences from large and diverse alignments of RNA viruses.
Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution.
Furthermore, our work has implications for dating of divergence times of RNA viruses.
We don't have solutions for the latter, but feel that it is still important to point them out.
}

@@

\refb{But what I’m missing the most from the results and discussion sections of the manuscript, is clear advice on how to perform analyses on real data, what to look out for and how to potentially remedy the problem. The manuscript is too technical in that regard, and while assessing accuracy problems is important, a step-by-step approach for empirical data sets- to potentially/hopefully mitigate the problems mention in this manuscript - is even more important. It would be most helpful if the authors comment on the typical workflow, i.e. running ModelTest or something similar (PartitionFinder, automatic/included model selection in IQ-TREE, ...) and then sticking with the model found to infer the phylogeny. How often would one get in trouble with such an approach? And if problems arise, how severe are they? Could the clustering potentially be impacted or does the problem stay limited to certain branches (of certain length)? Or is the tree as a whole deemed older/younger than it really is, i.e. does the underestimation propagate through the tree all the way to the root or not?}

\response{
We hope that the revision makes now clear that phylogenetic inference is not the focus of the manuscript, but that we instead study how properties of sites can be learned from sequence diversity.
Such inference from large viral data sets is increasingly feasible as data set size increases and complements deep mutations scanning approaches.
While we identify problems for phylogenetic inference associated with strong variation model variation along the sequence, we don't have a solution.
The site specific models we estimate could serve as input for ParitionFinder or the "-MF+MERGE" feature of IQ-tree.
}

@@

\response{In essence, Fig 3 shows that branch length underestimation caused by extensive variation of rates and preferences is very hard to avoid.
As outlined above, we don't really have a good fix for that, but we investigate the causes in Fig 4.
That said, the main result of the paper is that site-specific frequencies can be accurately estimated.
The inability is estimating the length of long branches in RNA viruses is a serious concern for dating zoonosis and exploring the history of viruses causing measles, dengue, lassa etc.
This problem is clearly apparent when trying to date the common ancestor of SIVs.
Molecular clock approaches suggest an ancestor a few hundred years in the past, while viruses in island populations show that the common ancestor is much more ancient.
}

\refb{The manuscript deals with estimates (i.e. trees) that were obtained using IQ-TREE and FastTree (which produce unrooted trees and don’t employ the concept of a molecular clock), which were subsequently fed into TreeTime which employs the provided trees as is. This makes me wonder if the conclusions from the manuscript hold up in frameworks that i) employ a molecular clock directly in their estimations, and ii) do not employ a fixed topology but rather explore a distribution of trees as typically done in Bayesian phylogenetic inference. I’m not suggesting the authors do the same experiments in a Bayesian framework but their take on this could be interesting for the discussion section.}

\response{
We are afraid there is a misunderstanding here.
This manuscript deals with GTR models and the root doesn't matter.
We use root-to-tip distances in mid-point-rooted trees as a measure of divergence, but the root is no special place otherwise.
The only place where time comes in are the HIV examples, but we only use time as a proxy for evolutionary rate and don't do any molecular clock analysis.
All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}

### 2020-07-27 manuscript/ms.tex
more edits
more edits, polish

\begin{abstract}
Abstract: Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutrally, while others are highly conserved, allow only specific states or only change in concert with other sites.
One one hand, such constraints on sequence evolution are used to infer biological function, one the other hand they need to be accounted for in phylogenetic reconstruction.
Most commonly used evolutionary models try to account for this complexity by partitioning sites into a small number of discrete classes with different rates and/or preferences where appropriate complexity is determined by model selection procedures.
Here, we present an efficient algorithm to estimate more complex models that allow for different preferences at every site and explore the accuracy at which such models can be estimated from simulated data.
Our iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
We further explore the importance of site-specific preferences during estimation of branch length of phylogenetic trees and show that ignoring site-specific preferences results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
Analysis of site-specific HIV substitution models suggests near saturation of divergence after a few hundred years.
Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
\end{abstract}

@@

In addition to rate variation, different sites in a protein differ in the amino acids they allow, and constraints on the DNA or RNA sequence impose additional levels of selection.
Such variation can again be accounted for by modeling multiple categories with different preferences for amino acids or nucleotides to which sites can be assigned or which can be integrated out during phylogenetic inference \citep{Lartillot01062004,kainer_effects_2015,shapiro_choosing_2006}.

@@

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\hat{\eqp}_i^a$ scales inversely with the tree length, suggesting the accuracy is limited by the Poisson statistics of observable mutations.
    This scaling holds for different number of sequences per tree ($n\in [100,300,1000,3000]$) and different degrees of rate variation.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence.
	Analogous results for alphabets of size $q=20$ are shown in Fig.~\ref{fig:dressed_supp}.
}
	\label{fig:dressed}
\end{figure*}

@@

In our first analysis summarized in Fig.~\ref{fig:dressed} we quantify the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
As expected, the average squared error $\chi^2$ decreases with the substitution rate and the size of the data set.
The error is well predicted by the average number of substitutions per site, which increases both with data set size and with the substitution rate.
The data in Fig.~\ref{fig:dressed}A are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the squared deviation is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.

The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (colored lines) and $\alpha=3$ (grey lines).
The scaling behavior $\chi_2 \sim (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
In the latter case of strong rate variation, the average accuracy is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.

@@

Instead of using only the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest reduction of the error (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
The error now continues to decrease up to root-to-tip distances of about one.
At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence (yellow line).
Similar improvements are achieved by optimizing tree branch lengths along with the model.
These improvements are consistent with \citet{gascuel_darwinian_2020} in that joint estimation of ancestral states and model is possible for Yule trees in which tree length grows linearly with the number of tips while tree depth increases only logarithmically.
In contrast, using the frequencies of different sequence states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).

@@

\subsection*{Implications for branch length estimates in phylogenies}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if uniform equilibrium frequencies are assumed.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{(\eqp_i^a)}^2$, which is increasing sharply with more peaked preferences.
Models that don't account for site-specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent evolutionary rates \citet{wertheim_purifying_2011}.

@@

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.7\textwidth]{{../figures/model_deviation_n300}.pdf}
	\caption{{\bf Sensitivity of branch length estimates on model misspecification}.
	Panels A \& B show the relative error in total tree length when using a mixture model as defined in Eq.~\ref{eq:mixture_model} for branch length inference. Panel A shows this error as a function of the mixing fraction $\alpha$ for $\langle \mu \rangle = 0.2$. Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\alpha=1$.
	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
	The models assume an alphabet size $q=4$ (nucleotides).
	}
	\label{fig:model_deviation}
\end{figure*}

@@

\subsection*{Minor model deviations can have substantial effects on divergence estimates}
In the previous section, we have observed that model deviations, for example the error made during model inference, can result in substantial errors in branch length estimates.
To investigate this effect more systematically, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows:
For a mixing fraction $\alpha$, we constructed a model with rates
\begin{equation}
\label{eq:mixture_model}
	\mu^a = \alpha \langle \mu^a\rangle + (1-\alpha) \mu^a_{\mathrm{true}}
\end{equation}
where $\langle \mu^a\rangle$ is the average of the rate.
Mixtures of site-specific frequencies are constructed analogously.

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\alpha = 0.3$.
If the same preferences are used for every site ($\alpha=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above that using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.

@@

Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ should depend on the fitness difference $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
\begin{equation}
	f^a_{ij} = \frac{1-e^{-2s^a_{ij}}}{1-e^{-2Ns^a_{ij}}} \approx
	\begin{cases}
		2s^a_{ij} & s^a_{ij}>0 \\
		2s^a_{ij} e^{2Ns^a_{ij}} & s^a_{ij} < 0
	\end{cases}
\end{equation}
On longer time scales, the fixation rates $f^a_{ij}$ correspond to transitions rates of the site-specific GTR model $Q^a_{ij}$.
The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations \citep{neher_genetic_2013}.
Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $i$
\begin{equation}
	e^{2N s_i^a} = \frac{\Gamma_{in}}{\Gamma_{out}} = \frac{\sum_j \pi_i^a W_{ij}}{\sum_j W_{ij}\pi_j^a}
\end{equation}
The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the multiplicative factor $2N$ is unknown.

@@

We further explored how model choice affects the maximum likelihood estimates of branch lengths.
As has been reported before, branch lengths are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern_evolutionary_1998,hilton_modeling_2018}.
While such underestimation often has a moderate effect on the total length of a tree, it can result in substantial underestimation of root-to-tip distance that are dominated by a few long branches close to the root.
While using the correct model results in correct branch length estimates, joint inference of site-specific models and branch lengths is typically unable to recover the true branch lengths and substantial underestimation remains.
This points to parameter identifiability problems rooted in the similar effects that skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
This problems are distinct from, but related to, the inability to jointly infer ancestral states and rate parameters of evolving discrete states \citep{gascuel_darwinian_2020} as we can marginalize over the ancestral states here.

>>>>

\begin{abstract}
Abstract: Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutrally, while others are highly conserved, allow only specific states, or only change in concert with other sites.
One one hand, such constraints on sequence evolution can be to infer biological function, one the other hand they need to be accounted for in phylogenetic reconstruction.
Phylogenetic models often account for this complexity by partitioning sites into a small number of discrete classes with different rates and/or state preferences.
Appropriate model complexity is typically determined by model selection procedures.
Here, we present an efficient algorithm to estimate more complex models that allow for different preferences at every site and explore the accuracy at which such models can be estimated from simulated data.
Our iterative approximate maximum likelihood scheme uses information in the data efficiently and accurately estimates site-specific preferences from large data sets with moderately diverged sequences and known topology.
However, the joint estimation of site-specific rates, and site-specific preferences, and phylogenetic branch length can suffer from identifiability problems, while ingoring variation in preferences across sites results in branch length underestimates.
Site-specific preferences estimated from large HIV \emph{pol} alignments show qualitative concordance with intra-host estimates of fitness costs.
Analysis of these substitution models suggests near saturation of divergence after a few hundred years.
Such saturation can explain the inability to infer deep divergence times of HIV and SIVs using molecular clock approaches and time-dependent rate estimates.
\end{abstract}

@@

In addition to rate variation different sites in a protein differ in the amino acids they allow, different positions in codons experience different constraints, and secondary structure or protein binding to DNA or RNA imposes additional levels of selection.
Such variation can again be accounted for by modeling multiple categories with different preferences for amino acids or nucleotides to which sites can be assigned or which can be integrated out during phylogenetic inference \citep{Lartillot01062004,kainer_effects_2015,shapiro_choosing_2006}.

@@

\begin{figure*}[htb]
	\centering
	\includegraphics[width=0.48\textwidth]{{../figures/p_dist_vs_treelength_nuc_ratealpha1.5}.pdf}
	%\includegraphics[width=0.32\textwidth]{../figures/avg_rate_dressed.pdf}
	\includegraphics[width=0.48\textwidth]{{../figures/rate_correlation_dressed_nuc_ratealpha1.5}.pdf}
	\caption{{\bf Accuracy of iterative estimation of site-specific GTR models as a function of the expected number of state changes along the tree.} (A) Mean squared error of the inferred $\hat{\eqp}_i^a$ scales inversely with the tree length (indicated by the black line), suggesting the accuracy is limited by the Poisson statistics of observable mutations.
    This scaling holds for different number of sequences per tree ($n\in [100,300,1000,3000]$) and different degrees of rate variation.
	(B) The relative substitution rates are accurately inferred as soon as the typical site experiences several substitutions across the tree as quantified here as Pearson correlation coefficient between true and inferred rates. Regularization via pseudo-counts reduces over-fitting at low divergence.
	Analogous results for alphabets of size $q=20$ are shown in Fig.~\ref{fig:dressed_supp}.
}
	\label{fig:dressed}
\end{figure*}

@@

In our first analysis summarized in Fig.~\ref{fig:dressed} we quantified the accuracy at which $\eqp_i^a$ and $\mu^a$ can be estimated if ancestral sequences are known.
As expected, the average squared error $\chi^2$ decreases with the substitution rate and the size of the data set.
The error is well predicted by the average number of substitutions per site, which increases both with data set size and with the substitution rate.
The data in Fig.~\ref{fig:dressed}A are shown on double logarithmic scales, such that the approximately straight decrease of $\chi^2$ with slope -1 implies that the squared deviation is inversely proportional to the expected number of substitutions.
This scaling suggests that accuracy is limited by the inherent stochasticity of the evolutionary process and that the inference uses all available information efficiently.

The simulation data were generated by drawing site-specific rates from a Gamma distribution with $\alpha=1.5$ (blue lines) and $\alpha=3$ (orange lines).
The scaling behavior $\chi_2 \sim (\textrm{tree length})^{-1}$ is more evident for $\alpha=3$ than for $\alpha=1.5$.
In the latter case of strong rate variation, the average squared error $\chi^2$ is dominated by a small number of sites with low rates at which frequencies are estimated poorly and the average tree length does not capture behavior at these sites.

@@

Instead of reconstructing the most likely ancestral sequences, we can instead average over all possible ancestral states, which results in a modest reduction of the error (``Marginalized ancestral sequences'', purple line in Fig.~\ref{fig:reconstructed}).
More significant gains are made when iterating model inference and ancestral reconstruction using the inferred site-specific model (``Iterative model estimation'', brown line).
The error now continues to decrease up to root-to-tip distances of about one.
At this level of divergence, tree reconstruction starts becoming problematic and branch lengths deviate substantially from their true values.
Using the true tree instead of the reconstructed tree leads to continuous improvements of accuracy with increasing levels of divergence (yellow line).
Similar improvements are achieved by optimizing tree branch lengths along with the model.
These improvements are consistent with \citet{gascuel_darwinian_2020} in that joint estimation of ancestral states and model is possible for Yule trees in which tree length grows linearly with the number of tips while tree depth increases only logarithmically.
In contrast, using the frequencies of different sequence states in the alignment as estimates for $\eqp_i^a$ is much less accurate due to the correlation induced by shared ancestry (orange lines in Fig.~\ref{fig:dressed}A).

@@

\subsection*{Implications for branch length estimates in phylogenies}
As previously observed by various authors \citep{halpern_evolutionary_1998,hilton_modeling_2018}, sites with heavily skewed preferences for specific states result in underestimation of branch lengths if these skews are not modeled appropriately.
This is a straightforward consequence of the fact that the probability of observing the same state at random is $\sum_i{(\eqp_i^a)}^2$, which is increasing sharply with more peaked preferences.
Models that don't account for site-specific frequencies will take a large number of sites that agree between two sequences as evidence for their close evolutionary relationship, while it is simply a consequence of resampling the same states with high probability.
In other words, this underestimation is the result of incorrect models of saturation and is closely related to the observation that incomplete purifying selection can result in erroneous and apparently time dependent evolutionary rates \citet{wertheim_purifying_2011}.

@@

\begin{figure*}[tb]
	\centering
	\includegraphics[width=0.7\textwidth]{{../figures/model_deviation_n300}.pdf}
	\caption{{\bf Sensitivity of branch length estimates on model misspecification}.
	Panels A \& B show the relative error in total tree length when using a mixture model as defined in Eq.~\ref{eq:mixture_model} for branch length inference. Panel A shows this error as a function of the mixing fraction $\gamma$ for $\langle \mu \rangle = 0.2$. Panel B shows the error as a function of the evolutionary rate $\langle \mu \rangle$ for $\gamma=1$.
	The mixing is applied to the equilibrium frequencies $\eqp_i^a$, the rates $\mu^a$, or both.
	The models assume an alphabet size $q=4$ (nucleotides).
	}
	\label{fig:model_deviation}
\end{figure*}

@@

\subsection*{The effect of model misspecification on divergence estimates}
In the previous section, we have observed that model deviations, for example the error made during model inference, can result in substantial errors in branch length estimates.
To investigate this effect more systematically, we constructed mixtures of the true model and a model with flat $\eqp_i^a$ and/or $\mu^a$ as follows:
For a mixing fraction $\gamma$, we constructed a model with rates
\begin{equation}
\label{eq:mixture_model}
	\mu^a = \gamma \langle \mu^a\rangle + (1-\gamma) \mu^a_{\mathrm{true}}
\end{equation}
where $\langle \mu^a\rangle$ is the average of the rate.
Mixtures of site-specific frequencies are constructed analogously.

Fig.~\ref{fig:model_deviation}A shows how deviation from the true model affect relative branch length estimates.
Deviation in rate estimates result immediately in underestimated branch length, while substantial effects of too flat site-specific preferences only manifest themselves once deviations are of order $\gamma = 0.3$.
If the same preferences are used for every site ($\gamma=1$), however, deviations are substantial.
Deviations in rate and preferences are approximately additive.
These observations are consistent with the finding above that using inferred preferences and true substitution rates result in more accurate branch length estimates than inferring both rates and preferences.

@@

Simple population genetic models predict that the fixation probability $f_{ij}^a$ of a mutation from state $j$ to state $i$ at site $a$ should depend on the fitness difference $s^a_{ij}$ and the effective population size $N$ as \citep{kimura_diffusion_1964}
\begin{equation}
	f^a_{ij} = \frac{1-e^{-2s^a_{ij}}}{1-e^{-2Ns^a_{ij}}} \approx
	\begin{cases}
		2s^a_{ij} & s^a_{ij}>0 \\
		2s^a_{ij} e^{2Ns^a_{ij}} & s^a_{ij} < 0
	\end{cases}
\end{equation}
On longer time scales, the fixation rates $f^a_{ij}$ correspond to transitions rates of the site-specific GTR model $Q^a_{ij}$.
The effective population size $N$ plays the role of a coalescent time scale $T_c$ and in general has little to do with census population sizes, in particular in rapidly adapting populations \citep{neher_genetic_2013}.
Nevertheless, the logarithm of the ratio $\log f^a_{ij}/f^a_{ji} \approx 2Ns^a_{ij}$ is an interpretable quantity related to the average fitness difference between states on time scales longer than the population genetic scale $N\sim T_c$.
We generalize this notion to multiple states and define a fitness score of state $i$ at position $a$ as the ratio of rates into and out of state $i$
\begin{equation}
	e^{2N s_i^a} = \frac{\Gamma_{in}}{\Gamma_{out}} = \frac{\sum_j \pi_i^a W_{ij}}{\sum_j W_{ij}\pi_j^a}
\end{equation}
The logarithm of this ratio is expected to be proportional to the fitness difference between state $i$ and the average alternative state at site $a$, while the common multiplicative factor $2N$ is unknown.

@@

We further explored how model choice affects the maximum likelihood estimates of branch lengths.
As has been reported before, branch lengths are underestimated when variation in rate or preferences are not fully accounted for \citep{halpern_evolutionary_1998,hilton_modeling_2018}.
While such underestimation often has a moderate effect on the total length of a tree, it can result in substantial underestimation of root-to-tip distance that are dominated by a few long branches close to the root.
While using the correct model results in correct branch length estimates, joint inference of site-specific models and branch lengths is typically unable to recover the true branch lengths and substantial underestimation remains.
This points to parameter identifiability problems rooted in the similar effects that skewed equilibrium frequencies, low rates, and shorter branch length have on the likelihood of the data.
This problems are related to the inability to jointly infer ancestral states and rate parameters of evolving discrete states \citep{gascuel_darwinian_2020}.
Model and tree inference alone, however, does not require acurate reconstruction of ancestral states as these can be marginalized.
